A Linear Solution of the Linearity Challenge

### 2025-12-23 paper/introduction.tex
revised intro

One of the challenges in the benchmark, henceforth called \emph{linearity
challenge}, concerns the formalization of a \emph{minimal calculus of sessions}.
%
Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
are established formalisms for the static analysis of distributed programs based
on peer-to-peer communication channels. Every session type system revolves
around three key ideas: (1) session endpoints are \emph{linear resources} that
cannot be discarded or duplicated without compromising some safety and liveness
properties of a program; (2) the type of a session endpoint is \emph{updated}
after each use to reflect the state of the interaction it describes; (3) peer
session endpoints are meant to be used in complementary ways so as to guarantee
the absence of communication errors and, to some extent, progress of the
interaction.
%
The linearity challenge is motivated by the observation that the proper handling
of linear resources in a formalization often requires a large number of
auxiliary definitions and technical results that divert from the main problem
under study~\cite{CarboneEtAl24}. One of the alleged culprits of such complexity
is \emph{context splitting}, namely the operation that partitions a typing
context in such a way that the linear resources described therein end up in only
one of the partitions. This observation has led to the exploration of various
alternatives techniques including leftover typing~\cite{ZalakainDardha21}, the
use of linearity predicates~\cite{SanoKavanaghPientka23} and tagged
contexts~\cite{ZackonEtAl25}.

In this paper we attack the linearity challenge from a different perspective:
\emph{instead of describing new techniques that are effective for the
formalization of the calculus in the challenge, we describe a new calculus that
is easy to formalize with known techniques}.\Luca{dobbiamo dire perché lo
spirito della challenge è rispettato, e dunque giustificare il titolo
dell'articolo. Dire che il nostro non è un calcolo ad hoc, ma ha fondamenti
logici importanti, è completo, ed è strettamente in relazione con altri modelli
analoghi (CP e linear pi)}
%
More specifically, we present a calculus called \LCP whose type system coincides
with the proof system of \emph{classical linear logic} and that features
\emph{linear channels} instead of sessions. Linear channels are meant to be used
for a \emph{single} communication. Structured sessions involving the exchange of
several messages can be modeled in \LCP using \emph{explicit continuations}
according to the well-known encoding of (binary) sessions into the linear
$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP is a low-level calculus with a strong connection with linear logic
and in which sessions are encodable instead of being featured natively.
%
These features play an important role in taming the complexity of its
formalization.
%
The fact of working with a calculus based on linear logic prevents \emph{by
construction} the same (sequential) process to own both endpoints of a session.
This situation is clearly undesirable since it does not correspond to a useful
pattern of interaction (every meaningful session requires its endpoints to be
used by parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of polarities~\citep{GayHole05} or
by using different names connected by the same
binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Also, the fact of working with a calculus with \emph{linear} channels and
explicit continuations spares us the need to \emph{update} the type of channels
in typing contexts. When a channel is used it is effectively consumed, therefore
its type can be \emph{removed} from the typing context and the type of the
continuation channel is \emph{added} back to the typing context. As is turns
out, removing and adding types is easier than updating them. Even more so
considering that the type of continuation channels can be conveniently added
\emph{at the beginning} of a typing context, since such channels are fresh by
definition.

In the end, the formalization of \LCP is both the most complete (in terms of
features supported by the calculus) and the most streamlined (in terms of code
size) among the known formalizations of session/linear
calculi~\cite{GotoEtAl16,Thiemann19,CastroFerreiraYoshida20,CicconePadovani20,RouvoetPoulsenKrebbersVisser20,SanoKavanaghPientka23,JacobsBalzerKrebbers22,ZackonEtAl25,ZalakainDardha21}.
It is also one of only two formalizations that prove the deadlock freedom
property for a session calculus, and it does so with substantially less code
than the existing formalization~\cite{JacobsBalzerKrebbers22}.

>>>>

One of the challenges in the benchmark, henceforth called \emph{linearity
challenge}, concerns the formalization of a \emph{minimal calculus of sessions}.
%
Sessions and session types~\citep{Honda93,HondaVasconcelosKubo98,HuttelEtAl16}
are established formalisms for the static analysis of distributed programs based
on peer-to-peer communication channels. Every session type system revolves
around three key ideas: (1) session endpoints are \emph{linear resources} that
cannot be discarded or duplicated without compromising some safety and liveness
properties of a program; (2) the type of a session endpoint is \emph{updated}
after each use to reflect the state of the interaction it describes; (3) peer
session endpoints are meant to be used in complementary ways so as to guarantee
the absence of communication errors and, to some extent, progress of the
interaction.
%
The linearity challenge is motivated by the observation that the proper handling
of linear resources in a formalization often requires a large number of
auxiliary definitions and technical results that divert from the main problem
under study~\cite{CarboneEtAl24}. One of the alleged culprits of such complexity
is \emph{context splitting}, namely the operation that partitions a typing
context in such a way that the linear resources described therein end up in only
one of the partitions. This observation has led to the exploration of various
alternative techniques including leftover typing~\cite{ZalakainDardha21}, the
use of linearity predicates~\cite{SanoKavanaghPientka23} and tagged
contexts~\cite{ZackonEtAl25}.

In this paper we attack the linearity challenge from a different perspective:
\emph{instead of proposing techniques that make it easy to formalize the
calculus in the challenge, we propose a (relatively new) calculus that is easy
to formalize with the known techniques}.\Luca{dobbiamo dire perché lo spirito
della challenge è rispettato, e dunque giustificare il titolo dell'articolo.
Dire che il nostro non è un calcolo ad hoc, ma ha fondamenti logici importanti,
è completo, ed è strettamente in relazione con altri modelli analoghi (CP e
linear pi)}
%
More specifically, we describe a calculus called \LCP whose type system
coincides with the proof system of \emph{classical linear logic} and that
features \emph{linear channels} instead of sessions. Linear channels are meant
to be used for a \emph{single} communication. Structured sessions involving the
exchange of several messages can be modeled in \LCP using \emph{explicit
continuations} according to the well-known encoding of (binary) sessions into
the linear
$\pi$-calculus~\citep{KobayashiPierceTurner99,Kobayashi02b,DardhaGiachinoSangiorgi17}.
%
Overall, \LCP is a linear version of \CP~\cite{Wadler14} in which sessions are
encodable instead of being featured natively.
%
These features play an important role in taming the complexity of its
formalization.
%
Using a calculus based on linear logic prevents \emph{by construction} the same
(sequential) process to own both endpoints of a session. This situation is
clearly undesirable since it does not correspond to a useful pattern of
interaction (every meaningful session requires its endpoints to be used by
parallel processes) and is a potential source of deadlocks.
%
From the standpoint of the formalization, where the representation of channel
names is a primary design choice, it spares us the need to \emph{distinguish}
the two endpoints of a session, \eg by means of
polarities~\citep{GayHole05,ZalakainDardha21} or by using different names
connected by the same binder~\citep{Vasconcelos12,GayVasconcelos25}.
%
Using a calculus with \emph{linear} channels and explicit continuations spares
us the need to \emph{update} the type of channels in typing contexts. When a
channel is used it is effectively consumed, therefore its type can be
\emph{removed} from the typing context and the type of the continuation channel
is \emph{added} back to the typing context. As is turns out, removing and adding
types is easier than updating them. Even more so considering that the type of
continuation channels can be conveniently added \emph{at the beginning} of a
typing context, since such channels are fresh by definition.

The formalization of \LCP that we obtain is both the most complete (in terms of
features supported by the calculus) and the most streamlined (in terms of code
size) among the known formalizations of session/linear
calculi~\cite{GotoEtAl16,Thiemann19,CastroFerreiraYoshida20,CicconePadovani20,RouvoetPoulsenKrebbersVisser20,SanoKavanaghPientka23,JacobsBalzerKrebbers22,ZackonEtAl25,ZalakainDardha21}.
It is also one of only two formalizations that prove the deadlock freedom
property for a session calculus, and it does so with substantially less code
than the existing formalization~\cite{JacobsBalzerKrebbers22}.
